---
title: "input_data"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# activate library functions
```{r message=FALSE}

library(ggplot2)
library(ggpattern)
#library(ggpubr)
library(expss)
library(readr)
library(readxl)
library(writexl)
library(skimr)
library(tidyverse)  # includes: tidyr, dplyr, haven, readr, readxl, forcats, stringer, ggplot2. See all with tidyverse_packages()
library(knitr)
library(janitor)
#install.packages("openxlsx")
library(openxlsx)
# install.packages("shinyaddins")
# library(shinyaddins)

```

```{r eval=FALSE}
# the purpose of this code is to input the stand data collected from the Arnot Forest (or other locations using the identical protocol).
# The code will create simple stand summaries, but no details analyses.
# The output summaries will be saved to a folder R/arnot_summary_output/filename(readme).xlsx
# 
# The original data are copied from BOX folder "Arnot Forest - Inventory Data and Projects/stand inventory data/data(entered)"
# The original data are presented in two worksheets, one "overstory" and one "understory"



### code 01
### Create file path input and the output folder (code 01)
```



file paths, input and output
```{r}


file_path_input <- "C:\\Users\\pjs23\\Documents\\R\\Arnot Stand Data\\rawdata\\arnot_inventory_data_01aug2025.xlsx"



output_folder <- "C:\\Users\\pjs23\\Documents\\R\\Arnot Stand Data\\summary_output\\"
```


```{r}

spp_type_lookup <- readr::read_csv(
  "C:/Users/pjs23/Documents/R/species and type table.csv",
  col_types = readr::cols(
    spp_code = readr::col_character(),
    spp_name = readr::col_character(),
    spp_type = readr::col_character()
  )
) %>%
  mutate(across(everything(), str_trim)) %>% 
    rename(
        spp = spp_code,
        type = spp_type
    )

# reads the table of spp numeric to alpha and "type"
# update the list of species codes within R/species table.csv

"C:\Users\pjs23\Documents\R\species and type table.csv"

spp_type <- read_csv("../species type table.csv", skip = 1, col_names = c("spp_code", "spp_name", "type")) 

spp_type_table <- tibble::tibble(
  spp = c(
    "ba","bc","bih","bl","co","he","ohw","rm","ro","sa",
    "ph","sh","sm","wa","wo","wp","yb","yp","comm_sdl",
    "sb",
    "ab","eh","pc","stm","toh","au",
    "app","bta","haw","brs","dws","ec","la","ns","cu","erc","ows","amc",
    "pb","gb","qa","rp","scp","smc","svb","wh","wi","viac","visp",
    "ah","cc","unk","sp","empty"
  ),
  type = c(
    rep("commercial", 19),
    "sb",
    rep("interfere", 6),
    rep("diversity", 22),
    rep("other", 5)
  )
)

```


```{r}
raw_overstory <- read_excel(file_path_input, sheet = "overstory", skip = 1) %>% 
    arrange(stand, growingseason, as.integer(point), spp)


overstory_prep <- raw_overstory%>% 
    mutate(
    baf = 9.1,
    point = as.character(point),
    growingseason = as.character(growingseason),
    spp = as.character(spp),
    status = as.character(status),
    crown = as.character(crown),
    stand = str_replace_all(stand, "-", "_")
    ) %>% 
    
#convert species numeric to species alpha
  left_join(spp_type_lookup, by = c("spp")) %>%
  mutate(
    spp = coalesce(spp_name, spp),
    spp = as.character(spp)
  ) %>%
  select(-spp_name) %>% 

    mutate(
    crown = case_when(
        crown == "0" ~ "dead",
        crown == "1" ~ "good",
        crown == "2" ~ "fair",
        crown == "3" ~ "poor",
        TRUE ~ crown  # keep other values unchanged
        ),
    status = case_when(
        status == "1" ~ "ags",
        status == "2" ~ "ugs",
        TRUE ~ status
    ),
    within16 = case_when(
      status %in% c("ugs", "dead") | dbh < 6 | dbh > 12 ~ 0,
      #(status != "ags") | (crown == "dead") & (dbh < 6 | dbh > 12) ~ 0,
      TRUE ~ within16
    ),
        status = if_else(crown == "dead", "dead", status)
    ) %>% 

    mutate(
        baf = 9.1,
        ba = baf,
        expansion = ba / (.005454 * dbh^2),  # (I'm not sure what this is doing) Maybe use this to back calculate the 9.1BAF value of 2.75*dbh=radius
        tpa = baf / ((0.005454 * dbh ^ 2)),
        reserve_tpa = (43560 / (pi * (16^2))) * within16 # the number of potential reserve trees per acre
    ) %>% 
    
    mutate(
        status = factor(status, levels = c("ags", "ugs", "dead") # to force arrange as ags, ugs, dead sequence
                        )) %>%


#    arrange(stand, as.numeric(point), spp, dbh)

     arrange(stand, spp, dbh, as.numeric(point)) #to assess the calculation of tree_count in df=overstory. confirmed 18aug2025

#   tree_count is the number of stems sampled by species as portrayed in df=stand_values


#need to calculate metrics for crown condition

overstory <- overstory_prep %>% 
    group_by(stand, point, spp, status) %>%  # values for each point
    summarize(
        type = first(type),
        growingseason = first(growingseason),
        tree_count = n(), # the number of trees measured during sampling
        sum_weighted = sum(dbh^2 * expansion),
        sum_treearea = sum(expansion),
        ba_sqft = sum(ba),
        stems_acre = sum(tpa),
        reserve_tpa = sum(reserve_tpa),
        .groups = "drop"
    ) %>% 
    mutate(
        quad_dbh = sqrt(sum_weighted/sum_treearea),
        dbhq2 = sqrt((ba_sqft/stems_acre)/0.005454)  # this is the same value as the var="quad_dbh"
    ) %>% 
    arrange(stand, as.numeric(point), spp, tree_count, stems_acre)


table(overstory$stand, overstory$spp)
#the numbers in the table are the number of points where the species is present
```

#create stand tables
```{r}
stand_table <- overstory_prep %>% 
  select(stand, point, growingseason, spp, dbh, tpa) %>% 
  mutate(
    # Bin dbh into 1-inch classes up to 30, then 30+
    dbh_class = case_when(
      dbh < 30 ~ paste0(floor(dbh), "_", floor(dbh) + 1),  # e.g., "10_11"
      TRUE     ~ "30plus"
    )
  ) %>%
  group_by(stand, growingseason, spp, dbh_class) %>% 
  summarize(
    tpa = sum(tpa),
    .groups = "drop"
  ) %>% 
  arrange(growingseason, stand, dbh_class) %>% 
  pivot_wider(
    id_cols = c(stand, growingseason, spp),
    names_from = dbh_class,
    names_glue = "dbh_{dbh_class}",
    values_from = tpa,
    values_fill = 0
  ) %>%
  # ---- reorder dbh columns numerically ----
  {
    dbh_cols <- grep("^dbh_", names(.), value = TRUE)
    # Extract the first number from the class (e.g., "dbh_10_11" -> 10, "dbh_30plus" -> 30)
    order_vals <- as.numeric(sub("dbh_(\\d+).*", "\\1", dbh_cols))
    ordered <- dbh_cols[order(order_vals)]
    dplyr::select(., growingseason, stand, spp, dplyr::all_of(ordered))
  } %>%
  # ---- change zeros to blanks ----
  mutate(across(starts_with("dbh_"), ~ ifelse(.x == 0, " ", as.character(round(.x, 0)))))
```


#write stand tables to unique worksheets
```{r}
# define file path
out_file <- file.path("C:\\Users\\pjs23\\Documents\\R\\arnot_forest_stand_analyses\\summary_output\\",
                      "stand_table.xlsx")

# create workbook
wb <- createWorkbook()

# loop through each stand and add as worksheet
unique_stands <- unique(stand_table$stand)

for (st in unique_stands) {
  # filter for the stand
  df_stand <- stand_table %>% filter(stand == st)
  
  # add worksheet with the stand name
  addWorksheet(wb, sheetName = paste0("stand_", st))
  
  # write data into that worksheet
  writeData(wb, sheet = paste0("stand_", st), x = df_stand)
}

# save workbook to your output folder
saveWorkbook(wb, out_file, overwrite = TRUE)

```



#calculate stand values by status (ags, ugs)
```{r}
n_points <- overstory %>%
  distinct(stand, point) %>%
  count(stand, name = "n_points")

stand_values01 <- overstory %>% 
    left_join(n_points, by = "stand") %>% 
    group_by(stand, growingseason, spp, status) %>%  # averages across points
    summarize(
        type = first(type),
 #       tree_count = sum(tree_count),
        ba_sqft_spp = sum(ba_sqft) / first(n_points),
        stems_acre_spp = sum (stems_acre) / first(n_points),
        quad_dbh_spp =  sqrt( (ba_sqft_spp / stems_acre_spp) / 0.005454),  #        dbhq2 = sqrt((ba_sqft/stems_acre)/0.005454) 

        .groups = "drop"
    ) %>% 
    mutate(
        file = "Arnot/input_data.Rmd",
        df = "stand_values01",
        size = "all"
    ) %>% 
    arrange(stand, status, spp) %>% 
    select(file, df, size, growingseason, stand, spp, status, type, ba_sqft_spp, stems_acre_spp, quad_dbh_spp) %>% 

    mutate_if(is.numeric, round, 1)

```

#calculate the stand-level (all stems) DBHq
```{r}
stand_values02 <- overstory %>% 
    group_by(stand, growingseason, type, point, status) %>% #average across species
    summarize(
#        type = first(type),
        ba_sum_pt = sum(ba_sqft),
        stems_acre_pt = sum(stems_acre),
        tree_count_pt = sum(tree_count),
        .groups = "drop"
    ) %>% 
    arrange(stand, growingseason, as.numeric(point)) %>% 
    
    left_join(n_points, by = "stand") %>% 
    
    group_by(stand, growingseason, type, status) %>% 
    summarize(
#        type = first(type),
        ba_acre_stand = sum(ba_sum_pt) / first(n_points),
        TPA_stand = sum(stems_acre_pt) / first(n_points),
        quad_dbh_stand =  sqrt( (ba_acre_stand / TPA_stand) / 0.005454),
        .groups = "drop"
    ) %>% 
    
    mutate_if(is.numeric, round, 1)
```

#calculate stand values for commercial species only

```{r}
stand_values_comm <- overstory %>% 
    filter(type == "commercial") %>% 
     group_by(stand, growingseason, point, status) %>% #average across species
    summarize(
        type = first(type),
        ba_sum_pt = sum(ba_sqft),
        stems_acre_pt = sum(stems_acre),
        tree_count_pt = sum(tree_count),
        .groups = "drop"
    ) %>% 
    arrange(stand, growingseason, as.numeric(point)) %>% 
    
    left_join(n_points, by = "stand") %>% 
    
    group_by(stand, growingseason, status) %>% 
    summarize(
        type = first(type),
        ba_acre_stand_comm = sum(ba_sum_pt) / first(n_points),
        TPA_stand_comm = sum(stems_acre_pt) / first(n_points),
        quad_dbh_stand_comm =  sqrt(((ba_acre_stand_comm) / (TPA_stand_comm)) / 0.005454),
        .groups = "drop"
    ) %>% 
     mutate_if(is.numeric, round, 1)

table(stand_values_comm$type)

```

#MERCHANTABLE
```{r}
#calculate stand values for merchantable species only
stand_values_merch <- overstory_prep %>% 
    filter(dbh > 12) %>% 

# code for "df = overstory"
    group_by(stand, point, type, status) %>%  # values for each point
    summarize(
        type = first(type),
        growingseason = first(growingseason),
        tree_count = n(), # the number of trees measured during sampling
        sum_weighted = sum(dbh^2 * expansion),
        sum_treearea = sum(expansion),
        ba_sqft = sum(ba),
        stems_acre = sum(tpa),
        reserve_tpa = sum(reserve_tpa),
        .groups = "drop"
    ) %>% 
    mutate(
        quad_dbh = sqrt(sum_weighted/sum_treearea),
        dbhq2 = sqrt((ba_sqft/stems_acre)/0.005454)  # this is the same value as the var="quad_dbh"
    ) %>% 

    group_by(stand, growingseason, point, type, status) %>% #average across species
    summarize(
        type = first(type),
        ba_sum_pt = sum(ba_sqft),
        stems_acre_pt = sum(stems_acre),
        tree_count_pt = sum(tree_count),
        .groups = "drop"
    ) %>% 

    left_join(n_points, by = "stand") %>% 
    
    group_by(stand, growingseason, type, status) %>% 
    summarize(
        ba_acre_stand_merch = sum(ba_sum_pt) / first(n_points),
        TPA_stand_merch = sum(stems_acre_pt) / first(n_points),
        quad_dbh_stand_merch =  sqrt(((ba_acre_stand_merch) / (TPA_stand_merch)) / 0.005454),
        .groups = "drop"
    ) %>% 
     mutate_if(is.numeric, round, 1)
```

# merge all sizes by type and merchantable by type
```{r}
stand_values <- stand_values02 %>% 
#    left_join(stand_values_comm, by = c("stand", "growingseason", "status")) %>% #commercial spp only
#    left_join(stand_values02, by = c("stand", "growingseason", "status")) %>%  #all stems, ignore species
    left_join(stand_values_merch, by = c("stand", "growingseason", "type", "status")) %>%  # all stems and spp >12" dbh
    mutate(
        file = "Arnot/input_data.Rmd",
        df = "stand_values",
        size = "all, merch >12"
    ) %>% 
    arrange(stand, status, type) %>% 
    select(file, df, size, growingseason, stand,  status, type, ba_acre_stand, TPA_stand, quad_dbh_stand,
           ba_acre_stand_merch, TPA_stand_merch, quad_dbh_stand_merch) %>% 
    mutate(across(where(is.numeric), ~replace_na(., 0))) %>% 
    mutate_if(is.numeric, round, 1)
    

    #stand_values01 %>% # all species and all sizes
    
# stand by spp by status all sizes = df_stand_values01
# stand by type by status ( all size & merch)  df=stand_values
    
    
table(overstory$stand, overstory$growingseason) #review stands by growing season
table(overstory$status, overstory$crown) # review status and crown. All status=dead should also be crown=dead
table(overstory$stand, overstory$spp)
table(overstory$within16, overstory$status) #run before executing the summarize code


```

#write data to individual worksheets, combine the spp data with the type data
```{r}
out_file2 <- file.path(
  "C:\\Users\\pjs23\\Documents\\R\\arnot_forest_stand_analyses\\summary_output\\",
  "stand_summary.xlsx"
)

common_stands <- intersect(stand_values01$stand, stand_values$stand)

wb <- createWorkbook()

for(s in common_stands) {
  
  addWorksheet(wb, sheetName = s)
  
  # Filter each dataset for this stand
  stand_values01_s <- stand_values01 %>% filter(stand == s)
  stand_values_s <- stand_values %>% filter(stand == s)
  
  # Ensure empty datasets are still included with column names
  if(nrow(stand_values01_s) == 0){
    stand_values01_s <- stand_values01_s[0,]  # empty df with same columns
  }
  if(nrow(stand_values_s) == 0){
    stand_values_s <- stand_values_s[0,]
  }
  
  # Combine data vertically with a source column
  combined <- bind_rows(
    stand_values01_s %>% mutate(source = "stand_values01"),
    stand_values_s %>% mutate(source = "stand_values")
  )
  
  # Write to Excel
  writeData(wb, sheet = s, combined)
}

saveWorkbook(wb, out_file2, overwrite = TRUE)











```



```{r}
understory <- read_excel(file_path_input, sheet = "understory", skip = 4) 

table(veg_data$harvest, veg_data$trmt, veg_data$forest_type)


```

